Sequent calculus style in Dedukti

### 2023-07-27 draft.tex
Update

	\subsection{Dedukti}

@@

@@

	In order to be able to declare lists over natural numbers, we need to define a type of natural numbers at the level of \textsf{Set} and map it to the one at the level of TYPE thanks to a rewriting rule.

@@

	\noindent Thanks to the induction principle over lists, it is possible to prove that \textcolor{blue}{$\Pi$} $\Gamma$, \textsf{append} $\Gamma$ \textsf{Nil} = \textsf{Nil}.

@@

	\subsection{First-Order Logic with Equality and Schematic Variables}
	We consider first-order logic with logical operators $(\neg, \land, \lor, \rightarrow, \leftrightarrow, \forall, \exists, \exists !)$, where $\exists !$ denotes the unique existential. We also consider $=$ as a built-in equality symbol, and accept schematic function and predicate symbols. A schematic symbol is a second-order variable that cannot be bound. It can be substituted, but behaves otherwise like an uninterpreted function or predicate symbol.

@@

	\begin{lstlisting}[language=Dialekto]
constant symbol Term : TYPE;
constant symbol Formula : TYPE;
	\end{lstlisting}
	Our encoding imposes no requirements on the structure of terms. In practice,  for every function symbol $f$ of arity $n$ in the source language, the user needs to define a corresponding symbol:
	\begin{lstlisting}[language=Dialekto]
constant symbol f :  Term → ... → Term
	\end{lstlisting}
	and similarly for predicate symbols
	\begin{lstlisting}[language=Dialekto]
constant symbol p :  Term → ... → Formula
	\end{lstlisting}
	As we consider first order logic with equality and have special rules for it, we define the equality symbol:
	\begin{lstlisting}[language=Dialekto]
constant symbol Equality :  Term → Term → Formula
	\end{lstlisting}
	And logical operators:
	\begin{lstlisting}[language=Dialekto]
symbol Neg : Formula → Formula;
symbol ∧ : Formula → Formula → Formula; notation ∧ infix right 30;
symbol ∨ : Formula → Formula → Formula; notation ∨ infix right 30;
symbol Implies : Formula → Formula → Formula;
symbol Iff : Formula → Formula → Formula;
\end{lstlisting}
%symbol Forall : (Term → Formula) → Formula;
%symbol Exists : (Term → Formula) → Formula;
%symbol ExistsOne : (Term → Formula) → Formula;
%	\end{lstlisting}
	\subsection{Higher-Order Abstract Syntax}
	Higher-Order Abstract Syntax (HOAS) is an approach to make shallow encoding of a system (in our case FOL) in an other (Dedukti) where variables of the original are represented as variable of the target system and not as syntactic objects. As it is not possible to express globally free variables in Dedukti, a formula with a free variable is encoded the following way:
	$$
	P(x_1) \land Q(x_1) \rightsquigarrow \lambda x_1 ,  P(x_1) \land Q(x_1)
	$$
	Note that the term on the right has type $\Term \rightarrow \Formula$. We then want
	$$
	\forall x_1, P(x_1) \land Q(x_1) \rightsquigarrow \forall (\lambda x_1 ,  P(x_1) \land Q(x_1))
	$$
	so that the universal quantifier $\forall$ is an operator with type $(\Term \rightarrow \Formula) \rightarrow \Formula$, and similarly for $\exists$ and $\exists !$. In Dedukti, we hence define:
		\begin{lstlisting}[language=Dialekto]
symbol Forall : (Term → Formula) → Formula;
symbol Exists : (Term → Formula) → Formula;
symbol ExistsOne : (Term → Formula) → Formula;
\end{lstlisting}
%	\end{lstlisting}
This type of encoding automatically takes care of, for example, capture-avoiding substitutions. Schematic symbols can be represented the same way. Consider for example the induction schema for natural numbers:
$$
	'P(0) \land (\forall n, 'P(n) \rightarrow 'P(n+1)) \rightarrow \forall n, 'P(n)
$$
It holds for any particular substitution of $'P$ by a formula with a distinguished variable. This is naturally represented in Dedukti by using abstracting over $'P$ with $\lambda 'P: \Term\rightarrow\Formula, ...$.

@@

	\fi
	\subsection{Sequent calculus}
	The Sequent Calculus LK is a proof system for first-order logic. A \textit{sequent} is a pair of ordered lists of formulas, noted
	$$
	\Gamma \vdash \Delta
	$$
	We usually use capital greek letter to denote sets of formulas. In an expression such as $\Gamma, \Sigma \vdash \phi, \Delta$, the comma denotes concatenation.
	The \textit{proof steps} of Sequent Calculus are depicted in \autoref{fig:scsteps1}, \autoref{fig:scsteps2} and \autoref{fig:scsteps3}. There are other formalizations of Sequent Calculus, including some where $\Gamma$ and $\Delta$ are \textit{sets} rather than list,as justified by the \texttt{Contract} and \texttt{Swap} rules, but this formalism is best suited for encoding in Dedukti and is the one we assume in this paper.

@@

	\section{Encoding of Sequent calculus in Dedukti}

>>>>

	This section introduces the logical framework Dedukti (Section~\ref{SSect:Dedukti}), the first-order logic (Section~\ref{SSect:FOL})

	higher-order abstract syntax (Section~\ref{SSect:HOAS})

	Sequent Calculus (Section~\ref{SSect:sequent-calculus}).

	\subsection{Dedukti}\label{SSect:Dedukti}

@@

	To define polymorphic lists
	As the declaration \lstinline[language=Dialekto, firstnumber=last]|constant symbol List: TYPE → TYPE;| is ill-formed since  a symbol with an  type.

@@

	In order to be able to declare lists over natural numbers, we need to define a type of natural numbers at the level of \textsf{Set} and map it to the one at the level of \lstinline[language=Dialekto, firstnumber=last]|TYPE| thanks to a rewriting rule.

@@

	\noindent Thanks to the induction principle over lists, it is possible to prove that \lstinline[language=Dialekto, firstnumber=last]|Π Γ, append Γ Nil = Γ|.

@@

	\subsection{First-Order Logic with Equality and Schematic Variables}\label{SSect:FOL}
	We consider first-order logic with logical operators $(\neg, \land, \lor, \rightarrow, \leftrightarrow, \forall, \exists, \exists !)$, where $\exists !$ denotes the unique existential. We also consider $=$ as a built-in equality symbol, and accept schematic function and predicate symbols. A schematic symbol is a second-order variable that cannot be bound. It can be substituted, but behaves otherwise like an uninterpreted function or predicate symbol.

@@

	\begin{lstlisting}[language=Dialekto]
constant symbol Term : TYPE;
constant symbol Formula : TYPE;
	\end{lstlisting}
	Our encoding imposes no requirements on the structure of terms. In practice,  for every function symbol $f$ of arity $n$ in the source language, the user needs to define a corresponding symbol:
	\begin{lstlisting}[language=Dialekto]
constant symbol f :  Term → ... → Term
	\end{lstlisting}
	and similarly for predicate symbols
	\begin{lstlisting}[language=Dialekto]
constant symbol p :  Term → ... → Formula
	\end{lstlisting}
	As we consider first order logic with equality and have special rules for it, we define the equality symbol:
	\begin{lstlisting}[language=Dialekto]
constant symbol Equality :  Term → Term → Formula
	\end{lstlisting}
	And logical operators:
	\begin{lstlisting}[language=Dialekto]
symbol Neg : Formula → Formula;
symbol ∧ : Formula → Formula → Formula; notation ∧ infix right 30;
symbol ∨ : Formula → Formula → Formula; notation ∨ infix right 30;
symbol Implies : Formula → Formula → Formula;
symbol Iff : Formula → Formula → Formula;
\end{lstlisting}
%symbol Forall : (Term → Formula) → Formula;
%symbol Exists : (Term → Formula) → Formula;
%symbol ExistsOne : (Term → Formula) → Formula;
%	\end{lstlisting}
	\subsection{Higher-Order Abstract Syntax}\label{SSect:HOAS}
	Higher-Order Abstract Syntax (HOAS) is an approach to make shallow encoding of a system (in our case FOL) in an other (Dedukti) where variables of the original are represented as variable of the target system and not as syntactic objects. As it is not possible to express globally free variables in Dedukti, a formula with a free variable is encoded the following way:
	$$
	P(x_1) \land Q(x_1) \rightsquigarrow \lambda x_1 ,  P(x_1) \land Q(x_1)
	$$
	Note that the term on the right has type $\Term \rightarrow \Formula$. We then want
	$$
	\forall x_1, P(x_1) \land Q(x_1) \rightsquigarrow \forall (\lambda x_1 ,  P(x_1) \land Q(x_1))
	$$
	so that the universal quantifier $\forall$ is an operator with type $(\Term \rightarrow \Formula) \rightarrow \Formula$, and similarly for $\exists$ and $\exists !$. In Dedukti, we hence define:
		\begin{lstlisting}[language=Dialekto]
symbol Forall : (Term → Formula) → Formula;
symbol Exists : (Term → Formula) → Formula;
symbol ExistsOne : (Term → Formula) → Formula;
\end{lstlisting}
%	\end{lstlisting}
This type of encoding automatically takes care of, for example, capture-avoiding substitutions. Schematic symbols can be represented the same way. Consider for example the induction schema for natural numbers:
$$
	'P(0) \land (\forall n, 'P(n) \rightarrow 'P(n+1)) \rightarrow \forall n, 'P(n)
$$
It holds for any particular substitution of $'P$ by a formula with a distinguished variable. This is naturally represented in Dedukti by using abstracting over $'P$ with $\lambda 'P: \Term\rightarrow\Formula, ...$.

@@

	\fi
	\subsection{Sequent calculus}\label{SSect:sequent-calculus}
	The Sequent Calculus LK is a proof system for first-order logic. A \textit{sequent} is a pair of ordered lists of formulas, noted
	$$
	\Gamma \vdash \Delta
	$$
	We usually use capital greek letter to denote sets of formulas. In an expression such as $\Gamma, \Sigma \vdash \phi, \Delta$, the comma denotes concatenation.
	The \textit{proof steps} of Sequent Calculus are depicted in \autoref{fig:scsteps1}, \autoref{fig:scsteps2} and \autoref{fig:scsteps3}. There are other formalizations of Sequent Calculus, including some where $\Gamma$ and $\Delta$ are \textit{sets} rather than list,as justified by the \texttt{Contract} and \texttt{Swap} rules, but this formalism is best suited for encoding in Dedukti and is the one we assume in this paper.

@@

	\section{Encoding of Sequent calculus in Dedukti}\label{Sect:SC-in-dedukti}
